Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions

### 2021-02-19 Latex/appendix/webappendix.Rnw
Updating text

\subsection{Normalization for non-linear biases via model offsets}

@@

Overall, we did not observe a significant improvement in performance by including input control in differential peak calling. Although several methods do offer the option of including controls in their analysis pipeline, we did not find that their inclusion was justifiable under our modelling assumptions. Our findings are in agreement with \cite{lun2015csaw}.

@@

\subsubsection{Simulation Study}

@@

@@

@@

\section{Bayesian Information Criterion (BIC) for Hidden Markov Models}

@@

\begin{figure}[h]
\begin{center}
\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_BIC/H3K36me3_3G_4R_1e+05W_100SNR_bic.png}}
\end{center}
\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 4 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
\label{fig:bic_4rep}}
\end{figure}

\begin{figure}[h]
\begin{center}
\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_BIC/H3K36me3_3G_2R_1e+05W_100SNR_bic.png}}
\end{center}
\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 2 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
\label{fig:bic_2rep}}
\end{figure}

\begin{figure}[h]
\begin{center}
\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_BIC/H3K36me3_3G_1R_1e+05W_100SNR_bic.png}}
\end{center}
\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 1 replicate per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
\label{fig:bic_1rep}}
\end{figure}

>>>>

\subsection{Normalization for non-linear biases via model offsets}
\label{s:loess}

@@

Overall, we did not observe a significant improvement in performance by including input control in differential peak calling. Although several methods do offer the option of including controls in their analysis pipeline, we did not find that their inclusion was justifiable under our modeling assumptions. Our findings are in agreement with \cite{lun2015csaw}.

@@

We performed a simulation study to assess the benefits and ability of epigraHMM to adjust for GC-content bias. In this study, we borrowed the method from \cite{teng2017accounting} and used their normalizing denominators to adjust for GC-content bias in our simulations study. As \cite{teng2017accounting} suggest, we utilized their normalizing offsets into our own model to account for the bias results from the simulated GC content. Because the simulated data varied with respect to the sequencing depth, we further included in our model normalizing offsets to adjust for the excess of read enrichment using the loess normalization approach as described in Section \ref{s:loess}.

Data was generated as described in Section 4.1 of our paper, with the addition that GC-content bias was introduced to the mean model. As suggested by \cite{teng2017accounting}, the simulated GC-content effect varied from condition to condition as well as the sequencing depth. In this simulation study, we assessed decreasing, neutral, and increasing effects of GC-content on the mean read count distribution. In Figure \ref{fig:sim_gc_scatter} we show an example of simulated data with increasing GC-content effect on ChIP-seq counts in both conditions. In this example, these conditions varied regarding the depth of sequencing reads.

@@

We assessed the model performance while adjusting for sequencing depth only, GC-content effect only, and simultaneously adjusting for both GC-content and sequencing depth. We observed that the optimal results were achieve only after normalization for both GC and sequencing depth bias. Despite the improvement in performance after the non-linear normalization, we were able to achieve optimal results in terms of true positive rate and false discovery rate of peak calls only after accounting for GC-content bias. These results not only show that our model is able to account for such type of bias, here computed using the methodology from \cite{teng2017accounting}, but also highlight the importance of such an adjustment. These results are shown in Figure \ref{fig:sim_gc_fdr}.

@@

To better visualize the results on a per data set basis, Figure \ref{fig:sim_gc_roc} shows the sensitivity and 1-specificity results from all 100 simulated datasets for the aforementioned example. Here, we see that for all data sets only after account for GC-content bias we were able to achieve optimal results.

@@

\section{Bayesian Information Criterion (BIC) for Hidden Markov Models}
\label{s:bicsec}

@@

We assessed the performance of our model selection scheme via BIC with a simulation study. To this end, we re-generated data in the same fashion as described in Section 4 of our paper while removing certain combinatorial patterns. Specifically, to mimic the ENCODE data analyzed in Section 5 of our paper, we generated data for which 3 conditions were available but only 2 out of 6 possible differential combinatorial patterns were available.

First, it is important to note that for an analysis of 3 conditions, there are $2^3-2 = 6$ possible combinatorial patterns of differential enrichment. In this context, for a mixture model with $L$ components, there are ${6\choose L}$ possible ways to model $L$ components. The choice of the number $L$ of components therefore needs to reflect the relevant biological patterns in the data. In this study, we assessed the algorithm presented in Section \ref{s:bicsec} by fitting all $\sum_{l=1}^{6}{6\choose l}$ possible model and computing their BIC for 100 simulated data sets. On average, the optimal chosen model should be the one with $L=2$ components \textit{and} with the components modeling those relevant ones. In this study, the two relevant patterns simulated were enrichment in condition 1 alone, and co-enrichment of conditions 2 and 3 (similar to Figure 6 of the main paper). As in Section 4 of our paper, we assessed the performance of our model under different SNR levels, different number of replicates per condition, and different genome sizes.

We present our results in Figures \ref{fig:bic_4rep}, \ref{fig:bic_2rep}, and \ref{fig:bic_1rep}. We start our model selection by fitting the full model with 6 components (L=6, rightmost block of the triangle with the highlighted '1'). We then proceed with our algorithm as described in this section and remove 1 combinatorial pattern at a time until the difference in BIC is large enough. As we can see from Figures \ref{fig:bic_4rep}, \ref{fig:bic_2rep}, and \ref{fig:bic_1rep}, this is achieved when L=2 and the combinatorial patterns being modeled is 1 (enrichment condition 1) and 2/3 (co-enrichment of condition 2 and 3). As we can see, our model selection scheme correctly selected the optimal model with the proper number of combinatorial patterns $L$ and the proper patterns themselves.

In accordance with the results presented in Section 4 of our paper, we note that the number of replicates played a major role in the success of our model selections scheme. For example, by comparing Figure \ref{fig:bic_4rep} and Figure \ref{fig:bic_1rep}, we see that more technical/biological replicates lead to models that can be better distinguished in terms of BIC when one is searching for the optimal patterns and number of combinatorial patterns. In summary, the presented model selection scheme allows users to prune combinatorial patterns of interest when one is not interested in modeling certain rare biological patterns from the data.

\begin{figure}[h]
\begin{center}
\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_BIC/H3K36me3_3G_4R_1e+05W_100SNR_bic.png}}
\end{center}
\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 4 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR. Dashed lines indicate order of model selection scheme beginning with the full model ('1') and ending with the reduced model ('6'). The optimal model is the one with highlighted '2', since it had a significant lower BIC than any other model with a single differential component. Highlighted number in each block indicated the combinatorial pattern being modeled. For instance '1/2 2 3' indicates a model with three components and the following patterns: co-enrichment of condition 1 and 2, enrichment of condition 2 alone, and enrichment of condition 3 alone. 
\label{fig:bic_4rep}}
\end{figure}

\begin{figure}[h]
\begin{center}
\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_BIC/H3K36me3_3G_2R_1e+05W_100SNR_bic.png}}
\end{center}
\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 2 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR. Dashed lines indicate order of model selection scheme beginning with the full model ('1') and ending with the reduced model ('6'). The optimal model is the one with highlighted '2', since it had a significant lower BIC than any other model with a single differential component. Highlighted number in each block indicated the combinatorial pattern being modeled. For instance '1/2 2 3' indicates a model with three components and the following patterns: co-enrichment of condition 1 and 2, enrichment of condition 2 alone, and enrichment of condition 3 alone. 
\label{fig:bic_2rep}}
\end{figure}

\begin{figure}[h]
\begin{center}
\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_BIC/H3K36me3_3G_1R_1e+05W_100SNR_bic.png}}
\end{center}
\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 1 replicate per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR. Dashed lines indicate order of model selection scheme beginning with the full model ('1') and ending with the reduced model ('6'). The optimal model is the one with highlighted '2', since it had a significant lower BIC than any other model with a single differential component. Highlighted number in each block indicated the combinatorial pattern being modeled. For instance '1/2 2 3' indicates a model with three components and the following patterns: co-enrichment of condition 1 and 2, enrichment of condition 2 alone, and enrichment of condition 3 alone. 
\label{fig:bic_1rep}}
\end{figure}

### 2021-02-19 Latex/paper/paper.tex
Updating text

First, we segmented the genome using the Viterbi sequence of most likely HMM states to understand the distribution of genomic regions associated with consensus background, differential, and consensus enrichment states (Figure~\ref{fig:fig_multimarks}). While the majority of the genomic regions exhibited no enrichment for any of the analyzed marks, regions of consensus enrichment were rare and represented only 2\% of the analyzed genome (Figure~\ref{fig:fig_multimarks}A), as expected. Consensus background and differential regions mostly covered intergenic (66\%) and protein-coding genic regions (61\%), respectively. Differential genomic windows were mostly representing either transcribed chromatin states or repressed chromatin states (Figure~\ref{fig:fig_multimarks}B). All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.05 and were pruned using our model selection scheme via BIC presented in Web Appendix B2 (Figure~\ref{fig:fig_multimarks}C). %. The Viterbi algorithm allows the simultaneous segmentation of the genome regarding the three HMM states without the necessity of choosing state-specific FDR control thresholds
\begin{figure}
\begin{center}
\includegraphics[scale = 0.7]{../../Main/Figure_MultiMarks/Figure_MultiMarks.pdf}
\end{center}
\caption{Genomic chromatin state segmentation and classification. (A): distribution of base pairs (y-axis) and the Viterbi sequence of states (x-axis). (B): estimated mixture probabilities and the associated differential combinatorial patterns. (C): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus. (D): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively. This figure appears in color in the electronic version of this article.
\label{fig:fig_multimarks}}
\end{figure}

@@

We presented a flexible and efficient statistical model designed to call differential regions of enrichment from ChIP-seq experiments with multiple replicates and multiple conditions. Our model has three main advantages over current methods tailored for differential peak detection. First, it uses an HMM-based approach that accounts for the local dependency of ChIP-seq counts and is able to precisely detect broad and short differential regions of enrichment. Second, it utilizes a GLM-based framework with model offsets that account for potential non-linear biases in the data as well as a constrained parametrization across HMM states and mixture components. Our implementation of the RCEM algorithm led to genome-wide analyses of ChIP-seq data under a computational time comparable to some of the fastest current methods and was at least 5 times faster than current HMM-based algorithms. Our results highlight the importance of the inclusion of technical/biological replicates in the analysis of epigenomic data. Lastly, our method allows the simultaneous detection and classification of differential combinatorial patterns of enrichment from its embedded mixture model and the associated posterior probabilities under any number of conditions. In situations where certain combinatorial patterns are rare, prunning of specific combinatorial patterns is possible via model selection via BIC (Web Appendix B2). Our software has been implemented into an R package and is available for download (Web Appendix A). % Data resulting from these assays are similar in nature but aim to identify regions that alternatively are not bound to proteins. % Such a tool could trigger, for instance, new insights on the field of epigenomics as investigators are often interested in making sense of the roles of epigenomic marks on the human genome. % Analysis of datasets from ther epigenomic assays, such as ATAC-seq, could also benefit from the presented method.

>>>>

First, we segmented the genome using the Viterbi sequence of most likely HMM states to understand the distribution of genomic regions associated with consensus background, differential, and consensus enrichment states (Figure~\ref{fig:fig_multimarks}). While the majority of the genomic regions exhibited no enrichment for any of the analyzed marks, regions of consensus enrichment were rare and represented only 2\% of the analyzed genome (Figure~\ref{fig:fig_multimarks}A), as expected. Consensus background and differential regions mostly covered intergenic (66\%) and protein-coding genic regions (61\%), respectively. Differential genomic windows were mostly representing either transcribed chromatin states or repressed chromatin states (Figure~\ref{fig:fig_multimarks}B). All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.05 and were pruned using our model selection scheme via BIC presented in Web Appendix B2 (Figure~\ref{fig:fig_multimarks}C). %. The Viterbi algorithm allows the simultaneous segmentation of the genome regarding the three HMM states without the necessity of choosing state-specific FDR control thresholds
\begin{figure}
\begin{center}
\includegraphics[scale = 0.7]{../../Main/Figure_MultiMarks/Figure_MultiMarks.pdf}
\end{center}
\caption{Genomic chromatin state segmentation and classification. (A): distribution of base pairs (y-axis) and the Viterbi sequence of states (x-axis). (B): estimated mixture probabilities and the associated differential combinatorial patterns. (C): choice of best model with two mixture components via BIC. (D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus. (D): association with differential H3K36me3 enrichment based based on mixture model posterior proportions of known marker genes. (E): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively. This figure appears in color in the electronic version of this article.
\label{fig:fig_multimarks}}
\end{figure}

@@

We presented a flexible and efficient statistical model designed to call differential regions of enrichment from ChIP-seq experiments with multiple replicates and multiple conditions. Our model has three main advantages over current methods tailored for differential peak detection. First, it uses an HMM-based approach that accounts for the local dependency of ChIP-seq counts and is able to precisely detect broad and short differential regions of enrichment. Second, it utilizes a GLM-based framework with model offsets that account for potential non-linear biases in the data as well as a constrained parametrization across HMM states and mixture components. Our implementation of the RCEM algorithm led to genome-wide analyses of ChIP-seq data under a computational time comparable to some of the fastest current methods and was at least 5 times faster than current HMM-based algorithms. Our results highlight the importance of the inclusion of technical/biological replicates in the analysis of epigenomic data. Lastly, our method allows the simultaneous detection and classification of differential combinatorial patterns of enrichment from its embedded mixture model and the associated posterior probabilities under any number of conditions. In situations where certain combinatorial patterns are rare, pruning of specific combinatorial patterns is possible via model selection via BIC (Web Appendix B2). Our software has been implemented into an R package and is available for download (Web Appendix A). % Data resulting from these assays are similar in nature but aim to identify regions that alternatively are not bound to proteins. % Such a tool could trigger, for instance, new insights on the field of epigenomics as investigators are often interested in making sense of the roles of epigenomic marks on the human genome. % Analysis of datasets from ther epigenomic assays, such as ATAC-seq, could also benefit from the presented method.
